---
title: "Basic Variable Usage"
output: pdf_document
date: "2024-06-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic Variable Usage

HD Sheets
created 6/19/2024

Practice Material for DSE 5002

#This is a tutorial

-
When looking at a tutorial, ask yourself:

    "Why is the author showing me this?"

    "What could I do with this idea?"
    
    "Imagine what would happen if you made a small change in the code.
       Then, try making that change and see if you were right.  If you
       were wrong, try to figure out why."

Try to remember the functions and operations used in tutorial, and where you 
saw them.

To get help use the help browser on the right, and use google as well.

#Creating some basic variables

R uses dynamic typing, so we don't have to tell R what type of variable each of
of variable is when we create it.

R has a set of rules it uses to do this that generally work well.

In statically typed languages, like Rust and C, you have to declare the 
variable type before you use it.

I'm using very simple data names here, more descriptive, detailed names are 
usually better

```{R}
a=5
b=4.0
c="Monday"
d=TRUE
```

#Action Required:

Ask yourself, what type or class of variable do I think the above variables
are?

Type your answers into the file here:



We can find out using the class() function

```{R}
class(a)
class(b)
class(c)
class(d)
```
Were there any surprises here?

We could also use the str() or structure function to learn about variables

```{R}

str(a)
str(b)
str(c)
str(d)

```

we can force variables to change type

```{R}
g=as.integer(a)
h=as.character(b)
```

Let's see what happened

```{r}
str(g)
str(h)
```
Did this produce the change you expected?


We can do math with our variables

```{R}

a+b
a*b
a^2

a/b

```
We can also make comparisons between variables

These result in logical variables, which are TRUE or FALSE

```{R}
a>b
a<2
a==5
a!=b
```
There are also the Boolean logic operators AND, OR and NOT
which are &,| and !

```{R}
TRUE|FALSE
TRUE&FALSE

(a<2) | (b>3)
(a!=b)&(b>4)
!d
```
#Lists

R can create lists of variables

These are sometimes called arrrays, in other languages, but are called lists 
in R

```{R}
list_a=c(1,3,5,7,9)
list_a
```
We can use both class() and str() to figure out what a list is

```{R}
class(list_a)
str(list_a)
```
We can put a list within a list, and this just yields a single, long list

```{R}
list_b=c( c(1,2), c(3,4),c(5,6))
list_b
```

What do you think happens if we try to put different types of objects in a list?

```{R}
list_c=c(1,2,3,"bob",5,6,"Wisconsin")
```

Action required

Enter some code to figure out what is in list_c and what type it is.
Is there one function that will do this for you?

```{R}

```

#Creating lists of sequences

There are a couple of ways of doing this

```{R}
list_x=1:10
list_x
list_y=seq(from=1,to=21, by=3)
list_y
```


#Vector operations

we can carry out operations on a list all at once

```{R}
list_c=3*list_a+4
list_c
```

Notice that are carried out the same operation on each item in the list

We can do this with logical operations or comparisons as well

```{R}
list_a<4

```
Notice that this gives us a list of true and false values, one for each entry 
in the list.

This ability will be very useful to us, as you will see in a moment

#Indexing a list or array

Indexing means to specify which items in a list we want

We can use a number to extract a specific entry in the list

```{R}
list_a
list_a[2]
```
Action required

Write code below to extract the 3rd item in list b

```{R}

```

We can extract several items at once

```{R}
list_a[c(1,3)]
```
#Indexing using an array of TRUE and FALSE values

We can also select items in an array by indexing them with a list of TRUE and
FALSE values the same length as the array

Only entries with a TRUE value will be extracted

```{R}

my_index=c(FALSE, TRUE, FALSE, FALSE,TRUE)
list_a
list_a[my_index]

```
#Slicing

Slicing an array means to select only the values that meet some specific 
condition

To do this we combine a test or comparision 

say list_a>6

which generates a list of TRUE/FALSE values with the indexing using TRUE/FALSE

Here it is

```{R}
list_a[(list_a)>6]

```
This maneuver, slicing using a test condition, is a really powerful move!

Remember this, you will use it over and over again.

It works in Python too!

#Other cool moves based on Logical Arrays

We can do some other moves with logical arrays as well

The any command checks to see if any of the items in a list meet the condition

```{R}

any((list_a==9))

```
 What do you think all() will do?
 
 Try it on list_b
 
 Find a condition that gives you a TRUE and another one that gives you a false
 
```{R}

```

Use the help function to look up the which() function

What does it do?

Try using it and a test condition on list_a

```{R}

```

Here's another odd one

Watch what happens if we sum() or add up a logical array

```{R}
g=list_a>4
g
sum(g)

```
What will sum( c(1,2,3,3,4)>2.1)  give us?

Predict the value, then try it

```{R}

```

#Sorting and ordering

We can sort a list in ascending or descending order

Note that this creates a new list

```{R}
list_x=1:10
list_x2=sort(list_x,decreasing=TRUE)
list_x
list_x2
```

Sometimes we will want the order in which to slice or arrange the list rather
than the sorted list itself.

This will become important when we start working with larger data structures

This is done with the order() function which will provide an index of the
list in sorted order

```{R}
list_k=c("a","q","x",4)
index_k=order(list_k,decreasing=FALSE)
list_k[index_k]

```





