---
title: "Basic Variable Usage"
output:
  html_document:
    df_print: paged
date: "2024-06-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Basic Variable Usage

HD Sheets
created 6/19/2024
updated 9/30/2024

# Practice Material for DSE 5002


Complete all the requested actions,  knit to a pdf and submit to the Pair 
Programming assignment for the week

# This is a tutorial

-
When looking at a tutorial, ask yourself:

    "Why is the author showing me this?"

    "What could I do with this idea?"
    
    "Imagine what would happen if you made a small change in the code.
       Then, try making that change and see if you were right.  If you
       were wrong, try to figure out why."

Try to remember the functions and operations used in tutorial, and where you 
saw them.

To get help use the help browser on the right, and use google as well.

# Creating some basic variables

R uses dynamic typing, so we don't have to tell R what type of variable each of
of variable is when we create it.

R has a set of rules it uses to do this that generally work well.

In statically typed languages, like Rust and C, you have to declare the 
variable type before you use it.

I'm using very simple data names here, more descriptive, detailed names are 
usually better

```{R}
a=5
b=4.0
c="Monday"
d=TRUE
```

# *Action Required:*

Ask yourself, what type or class of variable do I think the above variables (a,b,c,d)
are?

Type your answers into the file here:



We can find out using the class() function

```{R}
class(a)
class(b)
class(c)
class(d)
```
*Were there any surprises here?*

We could also use the str() or structure function to learn about variables

```{R}

str(a)
str(b)
str(c)
str(d)

```

# Changing Type

we can force variables to change type

```{R}
g=as.integer(a)
h=as.character(b)
```

Let's see what happened

```{r}
str(g)
str(h)
```
*Did this produce the change you expected?*

# Math

We can do math with our variables

These are the basic math operations on variables,  R can do all the basic operations as
well as linear algebra operations (coming later)

```{R}

a+b
a*b
a^2

a/b

```

#Comparisons

We can also make comparisons between variables

These result in logical variables, which are TRUE or FALSE

```{R}
a>b
a<2
a==5
a!=b
```

There are also the Boolean logic operators AND, OR and NOT
which are &,| and !

```{R}
TRUE|FALSE
TRUE&FALSE

(a<2) | (b>3)
(a!=b)&(b>4)
!d
```

# Vectors

R can create vectors of variables

All items in a vector must be the same type

We start the definition of a vector with c()

```{R}
vec_a=c(1,3,5,7,9)
vec_a
```

We can use both class() and str() to figure out what a array is

```{R}
class(vec_a)
str(vec_a)
```

We can put a vector within a vector,  creating one long vector

Note: In other languages,  this is called an array,  so if I use the term "array"
at points, think vector

```{R}
vec_b=c( c(1,2), c(3,4),c(5,6))
vec_b
```

What do you think happens if we try to put different types of objects in an vector?

```{R}
vec_c=c(1,2,3,"bob",5,6,"Wisconsin")
```

# *Action required*

Enter some code to figure out what is in list_c and what type it is.
Is there one function that will do this for you?

```{R}

```

#Creating vectors using automatic functions

There are a couple of ways of doing this

```{R}
vec_x=1:10
vec_x
vec_y=seq(from=1,to=21, by=3)
vec_y
```

A vector of random values

```{R}
vec_z=runif(5,0,1)
vec_z

```

#Vector operations

we can carry out operations on a vector all at once

```{R}
vec_c=3*vec_a+4
vec_c
```


Notice that are carried out the same operation on each item in the vector

We can do this with *logical operations* or *comparisons* as well

```{R}
vec_a<4

```

Notice that this gives us a list of true and false values, one for each entry 
in the list.

This ability will be very useful to us, as you will see in a moment

# Indexing a list or array

Indexing means to specify which items in a vector (or array) we want

We can use a number to extract a specific entry in the vector

```{R}
vec_a
vec_a[2]
```

# *Action required*

Write code below to extract the 3rd item in vector b

```{R}

```

We can extract several items at once

```{R}
vec_a[c(1,3)]
```

# Indexing using an array of TRUE and FALSE values

We can also select items in an array by indexing them with an array of TRUE and
FALSE values the same length as the array

Only entries with a TRUE value will be extracted

```{R}

my_index=c(FALSE, TRUE, FALSE, FALSE,TRUE)
vec_a
vec_a[my_index]

```
## Slicing

Slicing an array means to select only the values that meet some specific 
condition

To do this we combine a test or comparison 

say vec_a>6

which generates a vector of TRUE/FALSE values with the indexing using TRUE/FALSE

Here it is

```{R}
vec_a[(vec_a)>6]

```

This maneuver, slicing using a test condition, is a really powerful move!

Remember this, you will use it over and over again.

It works in Python too!

# Other cool moves based on Logical Arrays

We can do some other moves with logical arrays as well

The any command checks to see if any of the items in a list meet the condition

```{R}

any((vec_a==9))

```

# *Question/Action* 

 What do you think all() will do?
 
Try it on vec_b

Enter it in the cell below
 
```{R}
 
```
 
Find a condition that gives you a TRUE and another one that gives you a false
 
```{R}

```

#Question/Action

Use the help function to look up the which() function

What does it do?

Try using it and a test condition on list_a

```{R}

```

Here's another odd one

Watch what happens if we sum() or add up a logical array

```{R}
g=vec_a>4
g
sum(g)

```

# *Question/Action*

What will sum( c(1,2,3,3,4)>2.1)  give us?

Predict the value, then try it

```{R}

```

# Sorting and ordering

We can sort an vectotr in ascending or descending order

Note that this creates a new vector,   we need to store it somewhere

```{R}
vec_x=1:10
vec_x2=sort(vec_x,decreasing=TRUE)
vec_x
vec_x2
```

Sometimes we will want the order in which to slice or arrange the array rather
than the sorted array itself.

This will become important when we start working with larger data structures

This is done with the order() function which will provide an index of the
array in sorted order

```{R}
vec_k=c("a","q","x",4)
index_k=order(vec_k,decreasing=FALSE)
vec_k[index_k]

```

## *Question/Action*

Create a random vector of 10 values between 0 and 1 (see the earlier example on how to do this)

Then sort this vector in increasing order and save it to a new name

```{R}

```

## String variables

We can store strings (text) in r and we can create vectors of string variables

```{r}
a="Bob"
b="Smith"

d=c("alpha","beta","gamma")

a
b
d

```
We will see more later on how to work with text,  but the cat() function, 
short for concatenate, will paste strings together and print them

```{R}
cat(a,b,sep=" ")
```