---
title: "Week 1 Exercises"
author: "Your Name Here"
date: "May 9, 2022"
output: pdf_document
---

# Module 01 Lab 01   Homework assignment



Please complete all exercises below:



# Exercise 1

Assign 10 to the variable x. Assign 5 to the variable y. Assign 20 to the variable z.

```{r ex1}
#your code below

```


# Exercise 2

Show that x is less than z but greater than y.

**Note: your output must be a SINGLE boolean, do not output a boolean for each expression.**

```{r ex2}
#your code below

```

# Exercise 3

Show that x does not equal z and that y does not equal z

**Note: your output must be a SINGLE boolean, do not output a boolean for each expression.**

```{r ex3}
#your code here

```


# Exercise 4

Show that the formula `x + 2y = z`.  is true

**Note: your output must be a SINGLE boolean**

```{r ex4}
#your code below

```

# Exercise 5

I have created a vector (test_vector) of integers for you. Determine if any of x, y, or z are in the vector. 

**Note: your output must be a SINGLE boolean, do not output a boolean for each expression.**


```{r ex5}
test_vector <- c(1,5,11:22)
#your code below

```

# Exercise 6

Show which value is contained in the test vector. To do this you will need to create an element-wise logical vector using operators. `x == vector`. Once you have done that you will need to use slicing to return all indices that have matches.
**Note: your output should be two integers**

```{r ex6}
#your code below

```

#Problems from R for Data Science

https://r4ds.hadley.nz/workflow-basics

Exercises 2.5

1) Why does this code not work?

fix it so it works

```{R}
my_variable <- 10
my_varıable
```

Look carefully! (This may seem like an exercise in pointlessness, but training your brain to notice even the tiniest difference will pay off when programming.)


2.) Tweak each of the following R commands so that they run correctly:

```{R}
libary(todyverse)

ggplot(dTA = mpg) + 
  geom_point(maping = aes(x = displ y = hwy)) +
  geom_smooth(method = "lm)

```


3.) Press Option + Shift + K / Alt + Shift + K. What happens? How can you get to the same place using the menus?


4.) Let’s revisit an exercise from the Section 1.6. Run the following lines of code. Which of the two plots is saved as mpg-plot.png? Why?


```{R}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)

```


